### 2025-08-01 data-visualization.qmd
20250801 - tweaks

"Showing the data" involves showing the data that matters the most.
"Reducing the clutter" involves removing non-data things that obscure the data—for example, extraneous gridlines, tick marks, data markers (e.g., symbols to distinguish between series), and complex shadings (e.g., textured or filled gradients).
"Integrating the graphics and text" involves using headline titles, clear and useful labels (instead of legends), and helpful annotations.
Headline or newspaper-like titles are titles that are succinct with active phasing and that indicate the take-away message (e.g., "Quarterbacks Threw Fewer Touchdowns in 2024 than in Previous Years").
In terms of labels, @Schwabish2021 advocates to label the data directly instead of using a legend.
In terms of helpful annotations, you can provide additional text that helps explain the data (e.g., peaks or valleys, outliers, or other variations that deserve explanation), including how to interpret the chart.
"Avoiding the spaghetti chart" means avoiding packed charts with too much information that makes them difficult to interpret.
Spaghetti charts are lines with many lines that, make the plot look like a bunch of spaghetti.
However, @Schwabish2021 also advocates against using charts of other types that are complicated and difficult to interpret due to too much information, such as complicated maps or bar plots with too many colors, icons, or bars.
If there are too many lines or series, @Schwabish2021 advocates [breaking it up into multiple charts](#sec-faceting) (i.e., facets, trellis charts, or small multiples).
An example of [faceted charts](#sec-faceting) is depicted in @fig-facetedScatterplot.
"Starting with gray" refers to the idea of using gray as the default color for most lines/points/bars, so that you can use a color to highlight the lines/points/bars of interest.
In addition, as noted by @Schwabish2021, it is important to treat data as objectively as possible and not to present figures in a biased way as to mislead.

@@

The Data Visualization Catalogue provides examples of various types of plots depending on one's goal: <https://datavizcatalogue.com/search.html>.
The R Graph Gallery provides examples of various types of plots and how to create them in `R`: <https://r-graph-gallery.com>.
Books on data visualization in `R` include [*`ggplot2`: Elegant Graphics for Data Analysis*](https://ggplot2-book.org) [@ggplot3rdEd] and [*`R` Graphics Cookbook: Practical Recipes for Visualizing Data*](https://r-graphics.org) [@Chang2018].
In this chapter, we will examine how to create statistical graphics to visualize data.
We will create the plots using the `ggplot2` package [@R-ggplot2; @ggplot3rdEd].
When creating plots in `ggplot2` with multiple points or lines (e.g., multiple players or levels of a predictor variable), it is easiest to do so with the data in [long form](#sec-wideToLong) (as opposed to [wide form](#sec-longToWide)).

@@

The colors to use in the `fill` and `color` parameters can be specified by name (as in @fig-colorByName), hexadecimal (HEX) codes (as in @fig-colorByHEX), RGB values (as in @fig-colorByRGB), or by palettes (e.g., ColorBrewer or viridis; as in Figures [-@fig-multipleColorsByPaletteColorBrewer] and [-@fig-multipleColorsByPaletteViridis]).

@@

you can use any color by specifying its values in red-green-blue (RGB) color space using the `rgb()` function, as in @fig-colorByRGB.

```{r}
#| label: fig-colorByRGB
#| fig-cap: "Specifying the Color/Fill by RGB Value."
#| fig-alt: "Specifying the Color/Fill by RGB Value."

@@

```{r}
#| label: fig-multipleColorsByPaletteColorBrewer
#| fig-cap: "Specifying Multiple Colors Using the Color Brewer Palette."
#| fig-alt: "Specifying Multiple Colors Using the Color Brewer Palette."

>>>>

"Showing the data" involves showing the data that matter the most.
"Reducing the clutter" involves removing non-data things that obscure the data—for example, extraneous gridlines, tick marks, data markers (e.g., symbols to distinguish between series), and complex shadings (e.g., textured or filled gradients).
"Integrating the graphics and text" involves using headline titles, clear and useful labels (instead of legends), and helpful annotations.
Headline or newspaper-like titles are titles that are succinct with active phasing and that indicate the take-away message (e.g., "Quarterbacks Threw Fewer Touchdowns in 2024 than in Previous Years").
In terms of labels, @Schwabish2021 advocates to label the data directly instead of using a legend.
In terms of helpful annotations, you can provide additional text that helps explain the data (e.g., peaks or valleys, outliers, or other variations that deserve explanation), including how to interpret the chart.
"Avoiding the spaghetti chart" means avoiding packed charts with too much information that makes them difficult to interpret.
Spaghetti charts are lines with many lines that, make the plot look like a bunch of spaghetti.
However, @Schwabish2021 also advocates against using charts of other types that are complicated and difficult to interpret due to too much information, such as complicated maps or bar plots with too many colors, icons, or bars.
If there are too many lines or series, @Schwabish2021 advocates [breaking it up into multiple charts](#sec-faceting) (i.e., facets, trellis charts, or small multiples).
An example of [faceted charts](#sec-faceting) is depicted in @fig-facetedScatterplot.
"Starting with gray" refers to the idea of using gray as the default color for most lines/points/bars, so that you can use a color to highlight the lines/points/bars of interest.
In addition, as noted by @Schwabish2021, it is important to treat data as objectively as possible and not to present figures in a biased way as to mislead.

@@

The Data Visualization Catalogue, from Data to Viz, and Data Viz Project provide examples of various types of plots depending on one's goal: <https://datavizcatalogue.com/search.html>, <https://www.data-to-viz.com>, and <https://datavizproject.com>.
The R Graph Gallery provides examples of various types of plots and how to create them in `R`: <https://r-graph-gallery.com>.
Books on data visualization in `R` include [*`ggplot2`: Elegant Graphics for Data Analysis*](https://ggplot2-book.org) [@ggplot3rdEd] and [*`R` Graphics Cookbook: Practical Recipes for Visualizing Data*](https://r-graphics.org) [@Chang2018].
In this chapter, we will examine how to create statistical graphics to visualize data.
We will create the plots using the `ggplot2` package [@R-ggplot2; @ggplot3rdEd].
When creating plots in `ggplot2` with multiple points or lines (e.g., multiple players or levels of a predictor variable), it is easiest to do so with the data in [long form](#sec-wideToLong) (as opposed to [wide form](#sec-longToWide)).

@@

The colors to use in the `fill` and `color` parameters can be specified by name (as in @fig-colorByName), hexadecimal (HEX) codes (as in @fig-colorByHEX), values in red-green-blue (RGB) space (as in @fig-colorByRGB), or palettes (e.g., ColorBrewer or viridis; as in Figures [-@fig-multipleColorsByPaletteColorBrewer] and [-@fig-multipleColorsByPaletteViridis]).

@@

You can use any color by specifying its values in red-green-blue (RGB) color space using the `rgb()` function, as in @fig-colorByRGB.

```{r}
#| label: fig-colorByRGB
#| fig-cap: "Specifying the Color/Fill by Value in Red-Green-Blue (RGB) Space."
#| fig-alt: "Specifying the Color/Fill by Value in Red-Green-Blue (RGB) Space."

@@

```{r}
#| label: fig-multipleColorsByPaletteColorBrewer
#| fig-cap: "Specifying Multiple Colors Using the ColorBrewer Palette."
#| fig-alt: "Specifying Multiple Colors Using the ColorBrewer Palette."

### 2025-08-01 getting-started.qmd
20250801 - tweaks

1. Load the packages you need
1. Load the data files
1. Do any data processing needed to prepare the data files to be merged.
    - For example, make sure each data file has unique values on the keys.
1. Merge the data files
1. Do any data processing needed to prepare the data files for analysis.
For example:
    - calculate/recode variables
    - transform wide to long or long to wide
    - filter to the relevant rows
    - examine descriptive statistics to ensure data are plausible
- Do the analysis!
    - evaluate analysis assumptions
    - conduct sensitivity analyses to see extent to which findings differ using different subsets of the data, different modeling approaches/assumptions, etc.

>>>>

1. Load the packages you need
1. Load the data files
1. Do any data processing needed to prepare the data files to be merged.
    - For example, make sure each data file has unique values on the keys.
1. Merge the data files
1. Do any data processing needed to prepare the data files for analysis.
For example:
    - calculate/recode variables
    - transform wide to long or long to wide
    - filter to the relevant rows
    - examine descriptive statistics to ensure data are plausible
1. Do the analysis!
    - evaluate analysis assumptions
    - conduct sensitivity analyses to see extent to which findings differ using different subsets of the data, different modeling approaches/assumptions, etc.
